---
title: "Zarr arrays with Rarr" 
author: "Mike L. Smith"
output:   
  github_document:
    toc: true
    toc_depth: 2
---

<!-- badges: start -->
[![Codecov test coverage](https://codecov.io/gh/grimbough/Rarr/branch/devel/graph/badge.svg)](https://app.codecov.io/gh/grimbough/Rarr?branch=devel)
<br>
Build Status:
[![Bioconductor Release](https://bioconductor.org/shields/build/release/bioc/Rarr.svg)](https://bioconductor.org/checkResults/release/bioc-LATEST/Rarr/)
[![Bioconductor Devel](https://bioconductor.org/shields/build/devel/bioc/Rarr.svg)](https://bioconductor.org/checkResults/devel/bioc-LATEST/Rarr/)
<!-- badges: end -->

```{r, config, echo = FALSE}
knitr::opts_chunk$set(fig.path="inst/rmd/imgs/", dev = "jpeg")
```

# Introduction to Rarr

The Zarr specification defines a format for chunked, compressed, N-dimensional
arrays.  It's design allows efficient access to subsets of the stored array, and
supports both local and cloud storage systems. Zarr is experiencing increasing
adoption in a number of scientific fields, where multi-dimensional data are
prevalent. 

**Rarr** is intended to be a simple interface to reading and writing individual 
Zarr arrays. It is developed
in R and C with no reliance on external libraries or APIs for interfacing with
the Zarr arrays. Additional compression libraries (e.g. blosc) are bundled with
**Rarr** to provide support for datasets compressed using these tools.

## Limitations with **Rarr**

If you know about Zarr arrays already, you'll probably be aware they can be
stored in hierarchical groups, where additional meta data can explain the
relationship between the arrays.  Currently, **Rarr** is not designed to be
aware of these hierarchical Zarr array collections. However, the component
arrays can be read individually by providing the path to them directly.

Currently, there are also limitations on the Zarr datatypes that can be accessed
using **Rarr**.  For now most numeric types can be read into R, although in some
instances e.g. 64-bit integers there is potential for loss of information.
Writing is more limited with support only for datatypes that are supported
natively in R and only using the column-first representation.

# Quick start guide

```{r, child=c('inst/rmd/quick_start.Rmd')}
```


# Current Status

## Reading and Writing

Reading Zarr arrays is reasonably well supported. Writing is available, but is more limited.  Both aspects are under active development.  

### Data Types

Currently there is only support for reading and writing a subset of the possible datatypes
that can be found in a Zarr array.  In some instances there are also limitations on the 
datatypes natively supported by R, requiring conversion from the Zarr datatype.  The table below summarises the current status of
datatype support.  It will be updated as progress is made.

| Zarr Data Type | Status<br/>(reading / writing) | Notes |
|-----------|:--------------:|-------|
|`boolean`  | &#x2714; / &#x274C; | |
|`int8`     | &#x2714; / &#x274C; | |
|`uint8` | &#x2714; / &#x274C; | |
|`int16` | &#x2714; / &#x274C; | |
|`uint16`| &#x2714; / &#x274C; | |
|`int32` | &#x2714; / &#x2714; | |
|`uint32`| &#x2714; / &#x274C; | Values outside the range of `int32` are converted to `NA`.  Future plan is to allow conversion to `double` or use the [bit64](https://cran.r-project.org/package=bit64) package. | 
|`int64`|&#x2714; / &#x274C; | Values outside the range of `int32` are converted to `NA`. Future plan is to allow conversion to `double` or use the [bit64](https://cran.r-project.org/package=bit64) package. |
|`uint64`|&#x2714; / &#x274C; | Values outside the range of `int32` are converted to `NA`. Future plan is to allow conversion to `double` or use the [bit64](https://cran.r-project.org/package=bit64) package. |
|`half` / `float16`   |&#x2714; / &#x274C; | Converted to `double` in R.  No effort is made to assess loss of precision due to conversion.  |
|`single` / `float32` |&#x2714; / &#x274C; | Converted to `double` in R.  No effort is made to assess loss of precision due to conversion. |
|`double` / `float64` |&#x2714; / &#x2714; | |
|`complex`            |&#x274C; / &#x274C; | |
|`timedelta`          | &#x274C; / &#x274C; | |
|`datetime`           | &#x274C; / &#x274C; | |
|`string`             | &#x2714; / &#x2714; | |
|`Unicode`            | &#x274C; / &#x274C; | |
|`void *`             | &#x274C; / &#x274C; | |
| Structured data types | &#x274C; / &#x274C; | |

### Compression Tools

| Data Type   | Status<br/>(reading / writing) | Notes |
|-------------|:-------------------:|-------|
|`zlib / gzip`| &#x2714; / &#x2714; | Only system default compression level (normally 6) is enabled for writing. |
|`bzip2`      | &#x2714; / &#x2714; | Only compression level 9 is enabled for writing. |
|`blosc`      | &#x2714; / &#x2714; | Only `lz4` compression level 5 is enabled for writing. |
|`LZMA `      | &#x2714; / &#x2754; | |
|`LZ4`        | &#x2714; / &#x2714; | |
|`Zstd`       | &#x274C; / &#x274C; | Algorithm is available via blosc for writing, but can't currently be access through the R interface |

Please open an [issue](https://github.com/grimbough/Rarr/issues) if support for a required compression tool is missing.

### Filters

The is currently no support for additional filters.  Please open an [issue](https://github.com/grimbough/Rarr/issues) if you require filter support.


